---
title: "Cell Composition Report"
author: "Alfredo Rago"
date: "8/13/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggbiplot)
library(drake)
library(tidyverse)
library(here)
library(ggrepel)
library(magrittr)
```

## Preliminary data loading and re-formatting


```{r data_load}
# Normalized, annotated cell proportions from dtangle
loadd(tidyProps)
# Proportions annotaded with responder/non-responder groups
loadd(metaProps)
# Raw cell proportions from dtangle
loadd(cellPropRaw)

```

## Diagnostic plots

Checking what is the overall proportion of different cell types. Almost half of all cells are DC2 monocytes. This seems odd and could be indicative of issues with the markers.

```{r cellComposition, echo=FALSE}
group_by(cellPropRaw, ID) %>%
  summarize(., propMedian = median(Proportion)) %>%
  mutate(., cellType = fct_reorder(ID, -propMedian)) %>%
  ggplot(., aes(x = cellType, y = propMedian)) + 
  geom_col() +
  coord_flip()
```

Distribution within and between treatments are almost identical

```{r distribution, echo=FALSE}

ggplot(cellPropRaw, mapping = aes(x = Proportion, col = grepl("_L_", rowname))) +
  geom_density(alpha = 0.3) +
  facet_wrap(.~ID) +
  scale_x_log10()

```

There is a clear mean/variance relationship, which is expected in proportion data.


```{r variace_checks}
propVariance <- group_by(cellPropRaw, ID) %>%
  summarize(., 
            variance = var(Proportion),
            meanProp = mean(Proportion)
  ) %>%
  mutate(
    ID = factor(ID, ID[order(variance, decreasing = T)])
  )

ggplot(propVariance, aes(x = meanProp, y = variance, label = ID)) +
  geom_point() +
  geom_text_repel() +
  scale_y_log10() +
  scale_x_log10()
```

## Reaction norm plots

Visually check if any cell populations increase/decrease between sample and treatment across paired samples.

```{r reaction_norms_general, echo=FALSE}
 ggplot(data = tidyProps, aes(x = cond, y = proportion, group = subj)) +
  geom_line(aes(alpha=0.3)) +
  facet_wrap(.~cellType)
```

Now separated according to responder type

```{r reaction_norms_groups, echo=FALSE}
filter(metaProps, respond != "Outlier") %>% 
  mutate(., 
         # Replace whitespaces with newlines if they are preceeded and followed by long words
         cellType =  str_replace(cellType, "(?<=.{4,20}) (?=.{4,})", "\n"),
         # Placebo before LGG
         cond = factor(cond, levels = c("P", "L")),
         # Rename groups as in paper
         respond = factor(respond, levels = c("Group1", "Group2"), labels = c("LGG Responders", "LGG Non-Responders"))
           ) %>%  
  ggplot(data = ., aes(x = cond, y = proportion, group = subj, col = respond)) +
  geom_line(alpha=0.6) +
  scale_color_brewer(type = "qual", name = "Subject Group") + 
  facet_wrap(.~cellType) +
  theme_minimal() +
  scale_y_continuous(name = "Normalized Proportion") +
  scale_x_discrete(name = "") +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(),
        legend.position = c(0.8, 0.05),
        text = element_text(family = "serif", size = 16))

ggsave(filename = here("graphs/interactionPlot.pdf"),
      units = "mm", width = 297, height = 210)

ggsave(filename = here("graphs/interactionPlot.png"),
      units = "mm", width = 297, height = 210, dpi = 300)
```

And as a boxplot 

```{r boxplots}
ggplot(data = metaProps, aes(x = cond, y = proportion, group = respond, col = respond)) +
  geom_boxplot(position = 'dodge') +
  facet_wrap(.~cellType)

```

## Pseudo-MA plot

Immature goblet cells show a clear enrichment in LGG samples, so do WNT2 and CD4 but those have extremely low mean values.

```{r MAPlot, echo=FALSE}
set_colnames(cellPropRaw, c("sampleID", "cellType", "proportion")) %>%
  mutate(.,
         subj = str_extract(sampleID, pattern = "[0-9]{1,2}") %>%
           as.factor(),
         cond = str_replace_all(str_extract(sampleID, "_(L|P)_"), "_", "") %>%
           factor(levels = c('P', 'L')),
         cellType = as.factor(cellType)
  ) %>%
  select(., -"sampleID") %>%
  spread(., cond, proportion) %>%
  mutate(
    A = log2(L/P),
    M = (P+L)/2
  ) %>%
  group_by(cellType) %>%
  dplyr::summarise(
    P = mean(P),
    L = mean(L),
    A = mean(A),
    M = mean(M)
  ) %>%
  ggplot(., aes(x = M, y = A, label = cellType)) + 
  geom_point() + 
  scale_x_log10() +
  geom_label_repel()
  
```

## Statistical testing

Check if the cell type proportions change between placebo and LGG samples.
They do not.

```{r test}
metaProps_parallel <- select(metaProps, subj, cond, cellType, proportion) %>%
  spread(., 'cond', 'proportion') %>%
  group_by(., cellType)

general_responses <- summarise(metaProps_parallel, 
                               p.prop = wilcox.test(P, L, paired = T)$p.value,
                               ci.prop.min = wilcox.test(P, L, conf.int = T, paired = T)$conf.int[1],
                               ci.prop.max = wilcox.test(P, L, conf.int = T, paired = T)$conf.int[2]
)
general_responses$q.prop <- p.adjust(general_responses$p.prop, method = "fdr")

arrange(general_responses, q.prop)[1:5,]
```

## Check for differences in placebo cell composition

All the placebo samples have the same cell proportions for each type.
This means the biopsies of either group are similar.

```{r placebo_proportions}
placeboProps <- select(metaProps, -"week", -"sampleID") %>%
  filter(., cond == "P") %>%
  spread(., respond, proportion) %>%
  group_by(., cellType)

placeboPvals <- summarise(placeboProps, 
                          p.val = wilcox.test(Group1, Group2)$p.value
) %>% 
  mutate(., padj.val = p.adjust(p = p.val, method = "fdr"))

arrange(placeboPvals, padj.val)
```

## Check differences in LGG cell composition

LGG responder samples are identical to LGG non-responders in cell composition.
Glial cells are the most significant ones, with a 0.4 p.adj.
This means the changes in gene expression between responder and non-responder groups cannot be explained by changes in cell composition.

```{r LGG_proportions}
lggProps <- select(metaProps, -"week", -"sampleID") %>%
  filter(., cond == "L") %>%
  spread(., respond, proportion) %>%
  group_by(., cellType)

lggPvals <- summarise(lggProps, 
                      p.prop = wilcox.test(Group1, Group2)$p.value,
                      CV.lo = wilcox.test(Group1, Group2, conf.int = T)$conf.int[1] %>%
                        divide_by(., mean(mean(Group1, na.rm = T) + mean(Group2, na.rm = T))),
                      CV.hi = wilcox.test(Group1, Group2, conf.int = T)$conf.int[2] %>%
                        divide_by(., mean(mean(Group1, na.rm = T) + mean(Group2, na.rm = T)))
) %>% 
  mutate(., q.prop = p.adjust(p.prop, method = "fdr"))

arrange(lggPvals, q.prop)[1:5,]
```


## Delta version of paired test: focus on relative change

Analyze only differences between treatment and control (controls for different baselines).
Cell type composition is not significantly different between the two groups. 
Enteroendocrine cells are the most significant (qval = 0.1)
This means the differences between placebo and control groups cannot be explained by changes in cell composition.

```{R delta_tests}
deltaProps <- select(metaProps, -"week", -"sampleID") %>%
  spread(., cond, proportion) %>%
  mutate(
    delta = P-L,
    M = (P+L)/2
  ) %>%
  group_by(., cellType)

deltaPropsParallel = select(deltaProps, "subj", "cellType", "delta", "respond") %>%
  spread(., respond, delta) %>%
  group_by(., cellType)

deltaResponse <- summarise(deltaPropsParallel, 
                               p.prop = wilcox.test(Group1, Group2, paired = F)$p.value,
                               ci.prop.min = wilcox.test(Group1, Group2, paired = F, conf.int = T)$conf.int[1],
                               ci.prop.max = wilcox.test(Group1, Group2, paired = F, conf.int = T)$conf.int[2]
)

deltaResponse$q.prop <- p.adjust(deltaResponse$p.prop, method = "fdr")

arrange(deltaResponse, q.prop)[1:5,]
```


## PCA plot to visualize overall changes in composition

No overall pattern, although we can separate our responders using a combination of PC1 (low) and PC2 (low). This indicates that the response in group 1 is characterized by an increase in goblet cells and monocytes, and a decrease in stem and follicular cells.
In contrast, group 2 and outliers are scattered around the origin: this indicates no consistent change in cell composition.
Note that none of these changes is picked up by cell-by-cell testing, possibly because of low power.
We could run tests on the PCs if we suspect that the joint change in several cell types is important.

```{R PCA_deltas}
deltaPropsWide <- deltaProps[,c("cellType", "subj", "delta", "respond")] %>%
  spread(., cellType, delta) %>%
  column_to_rownames(., var = "subj")

deltaPCA <-  deltaPropsWide[, -1] %>%
  as.matrix(.) %>%
  prcomp(., center = T, scale. = T)

ggscreeplot(deltaPCA)

ggbiplot(deltaPCA, 
         choices = c(1,2),
         groups = deltaPropsWide$respond,
         ellipse = T)

ggbiplot(deltaPCA, 
         choices = c(1,3),
         groups = deltaPropsWide$respond,
         ellipse = T)

```
