---
title: "Cell Composition Report"
author: "Alfredo Rago"
date: "8/13/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

## Preliminary data loading and re-formatting


```{r data_load}
# Cell proportions from linseed dataset
load(file = here("/output/C4/linseedProportions.RData"))
# File with raw cell by sample proportions
cellProportions2 <- t(cellProportions) %>%
  as.data.frame(.) %>%
  rownames_to_column(.) %>% 
  as_tibble(.)
names(cellProportions2) <- str_replace_all(string = names(cellProportions2), pattern = " ", replacement = "")
# Tidy version: proportions annotated with sample ID and treatment/date
tidyprops <- gather(cellProportions2, key = "CellType", value = "prop", ... = -rowname) %>%
  mutate(.,
         subj = str_extract(rowname, pattern = "[0-9]{1,2}") %>%
           as.factor(),
         cond = str_replace_all(str_extract(rowname, "_(L|P)_"), "_", "") %>%
           factor(levels = c('P', 'L')),
         week = str_extract(rowname, pattern = "W[:digit:]") %>%
           factor()
  )

# Sample annotation from experiment
metadata <- read_tsv(file = here("../Data/Clinical_characteristics_Herlevstudy.txt")) %>%
  select(., contains("Screening"), contains("Treatment"), contains("Group")) %>%
  gather(., key = "Visit", value = "Treatment", ... = contains("Treatment"))
metadata$Visit <- str_extract(string = metadata$Visit, pattern = "[0-9]") %>% 
  paste0("W", .) %>%
  as.factor()
metadata$Treatment <- str_extract(string = metadata$Treatment, pattern = "^[A,P]") %>%
  str_replace(., "A", "L") %>%
  as.factor()
names(metadata) <- c('subj', 'respond', 'week', 'cond')

```

## Diagnostic plots

We check whether there are any cell populations that show odd distributions: i.e. specific to one sample only.

```{r diagnostic, echo=FALSE}
as.data.frame(cellProportions) %>%
  rownames_to_column() %>%
  gather(-1 ,key = 'ID', value = 'Proportion') %>%
  ggplot(., mapping = aes(x = Proportion, fill = rowname)) +
  geom_density(alpha = 0.3)
```

## Reaction norm plots

Visually check if any cell populations increase/decrease between sample and treatment across paired samples.

```{r reaction_norms_general, echo=FALSE}
 ggplot(data = tidyprops, aes(x = cond, y = prop, group = subj)) +
  geom_line(aes(alpha=0.3)) +
  facet_wrap(.~CellType)
```

Now separated according to responder type

```{r reaction_norms_groups, echo=FALSE}
metaprops <- merge(tidyprops, metadata, all.x = T, all.y = T) %>% 
  na.exclude() %>%
  as_tibble()

ggplot(data = metaprops, aes(x = cond, y = prop, group = subj)) +
  geom_line(aes(alpha=0.3)) +
  facet_grid(respond~CellType)
```

## Statistical testing

Check if the overall proportions change using a simple Wilcox rank sum test for paired samples

